Exploitation Techniques and Mitigations

### 2016-01-27 docs/paper.tex
Reorganize TikZ stuff and continue draft

@@

Exploiting binaries is an extremely broad topic with many specialized
techniques for even the most exotic scenarios. In this writeup we will take a
look at exploiting \texttt{printf} with crafted format strings first and follow
up with an introduction about buffer overflows. Mitigation mechanisms will be
disabled at first and enabled one by one discussing them when they are put into
place. The buffer overflow will be augmented to inject and execute shell code
which is then prevented by the Data Execution Prevention (DEP) mechanism.

Return Oriented Programming (ROP) is introduced together with ret2libc to
circumvent DEP. Address Space Layout Randomization (ASLR) is presented next as
counter to ROP but is quickly broken with an information leak. StackGuard is a
more sophisticated mechanism against ROP but not a silver bullet and can be
easily brute forced in certain scenarios. Control-Flow Integrity (CFI) together
with a word about Stack Integrity is provided as an outlook for the reader.

Some basic knowledge about the target platform (x86) is communicated before
running the first exploit, but we take a quick glance at other architectures
(x86\_64 and ARM) before concluding this writeup.

@@

Again the name already tells you what this is about, a direct acyclic graph
(DAG) that reflects the control-flow of your program. There are different
definitions regarding what is actually placed in the nodes of the graph and
what are the anchor points of the program used to connect them. In our case we
will create a node out of each function body and connect them at function
calls.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{gfx/unraid_part_graph.png}
    \caption{Truncated control-flow graph example}
    \label{fig:cfg}
\end{figure}

\Cref{fig:cfg} shows part of such a graph in
IDA\footnote{\url{https://www.hex-rays.com/products/ida/}} analyzing the
\texttt{emhttp} binary used by
unRAID\footnote{\url{http://lime-technology.com/what-is-unraid/}}. There are
much nicer graphs available to explain the concept, but we wanted to inject
some real world data. From the graph you can already tell that the function
(subroutine) labeled \texttt{licensing} will call \texttt{malloc},
\texttt{memset}, \texttt{USB\_STUFF}, \dots. Thats actually all we care about
for now.

@@

CFI is a big topic and, like other topics already mentioned, goes beyond the
scope of this writeup. The first pointer we will hand you aims at corresponding
section\footnote{\url{http://clang.llvm.org/docs/ControlFlowIntegrity.html}} of
the Clang documentation, but we encourage you to checkout the related research
paper \cite{ms_cfi} for more information. A more accessible and recent way to
this topic may be the
talk\footnote{\url{https://www.youtube.com/watch?v=FA0VK7s5tSQ}} \textit{New
memory corruption attacks: why can't we have nice things?} given by gannimo
(Mathias Payer).

The CFG has already been established, now let's see how it can be used to
counter the buffer overflow return address dilemma. At compile-time the graph
is available and can be used to create additional constraints which the program
must obey during runtime. This is similar to the StackGuard mechanism where we
attach code to the end of a function which checks if the canary is still
intact. But now we don't check for a canary but for the validity of the return
address. From the CFG we can build a set of possible return targets for each
function. Looking back at the example shown in \cref{fig:cfg} we can determine
that the function \texttt{szDecrypt} returns either to \texttt{USB\_STUFF},
\texttt{licensing} or one other subroutine excluded by the illustration. The
corresponding addresses are put into this set which is then stored in the
binary. Before the function \texttt{szDecrypt} returns the return address on
the stack is compared to the entries listed in the corresponding set. If no
match can be found, the program terminates.

With this mechanism setup, one can easily see that it is no longer possible
chain \emph{arbitrary} gadgets together to pull off ROP. But we still control
the return address and can jump to different locations as long as we stick to
the CFG. Each transition from one node to another has to be valid, while the
overall path taking by our ROP chain may do things never intended by the
program's author. This concept is known as control-flow bending.~
\cite{cf_bending}

@@

Using a changed return address is what ultimately enables control-flow bending.
Stack integrity ensures that the same return address is used upon executing the
\texttt{ret} instruction as was pushed upon function call. This can be achieved
by using a \textit{shadow-stack} similar to the StackShield mechanism but we
suggest reading about \textit{code-pointer integrity} \cite{cpi} regarding this
topic.

@@

We conclude this section by mentioning the availability of an interpreter
(probably) available in your standard library. As presented by gannimo,
\texttt{printf} is far more capable than just printing arguments. It is also
possible to read and \emph{write} to memory locations. But one can go even
further and craft a format string mimicking each of the eight operators of
Brainfuck\footnote{\url{https://en.wikipedia.org/wiki/Brainfuck}}. Because
Brainfuck is Turing complete we can deduce that \texttt{printf} is a Turing
complete interpreter. Note that this requires \texttt{printf} to be called in a
loop and the format strings may depend heavily on your libraries
implementation. (Also not all implementation are Turing complete).

@@

x86\_64, also known as x64 and AMD64, is at its core a \SI{64}{\bit} extension
to x86 which already replaced a lot of x86 machines. Its general purpose
registers are \SI{64}{\bit} wide and there are eight more of them.

@@

Contrary, breaking ASLR gets much harder since the address space is \emph{much}
bigger which yields more entropy for the randomization. This renders brute
force (and heap spraying) basically useless, but we can still resort to the
info leak. Breaking a canary via brute force gets only a little bit harder, but
this has already been shown.

@@

ARM CPUs will be encountered mostly in portable, low-power-oriented devices
such as smart phones and tables, but are also used in embedded devices like
routers. It consists of an \SI{32}{\bit} RISC instruction set with a
\SI{16}{\bit} mode (THUMB). The used calling convention is basically the same
as under x86\_64 (fastcall)

Compared to the previous both, ARM has a \SI{4}{\byte} instruction alignment
--- \SI{2}{\byte} under THUMB.

@@

String with no mitigation mechanisms in place, we have seen how easy it is to
manipulate the program by exploiting just one simple buffer overflow. Going
beyond simple manipulations like changing locale variables, we craft shell code
and injected it to open up a shell accepting and executing arbitrary inputs.
Next, Data Execution Prevention (DEP) was presented to deny the ability of
\emph{injecting} new code.

This was countered by introducing Return Oriented Programming (DEP) (and
ret2libc) which removes the requirement of injecting new code to exploit a
binary. This is done by combining code fragments (gadgets) already available in
the target binary and libraries to build new, malicious sequences of
instructions.

Address Space Layout Randomization (ASLR) can be defeated with an information
leak and even the StackGuard mechanism can broking with brute force (including
an unexpected low upper bound) in certain scenarios.

The basic idea behind Control-Flow Integrity (CFI) was communicated after this
followed by a small glance at \texttt{printf}'s capabilities to work as an
interpreter. Along the way references have been provided to aid the reader.

Before concluding with this section a short word about other architectures and
their influence on these techniques has been given,

>>>>

\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning,shapes.geometric}
\usetikzlibrary{shapes.multipart}

@@

Exploiting binaries is an extremely broad topic with many specialized
techniques for even the most exotic scenarios. In this writeup we will take a
narrow look at exploiting \texttt{printf} with crafted format strings first and
follow up with an introduction about buffer overflows. Mitigation mechanisms
will be disabled at first and enabled one by one --- discussing them when they
are put into place. The buffer overflow will be augmented to inject and execute
shell code which is then prevented by the Data Execution Prevention (DEP)
mechanism.

Return Oriented Programming (ROP) is introduced together with ret2libc to
circumvent DEP. Address Space Layout Randomization (ASLR) is presented next as
counter to ROP but gets quickly broken with an information leak. StackGuard is
a more sophisticated mechanism against ROP but not a silver bullet and can be
easily brute forced in certain scenarios. Control-Flow Integrity (CFI) together
with a word about Stack Integrity is provided as an outlook for the reader.

Although x86 has been chosen as target platform a quick glance at other
architectures (x86\_64 and ARM) is taken prior concluding this writeup. Some
basics about the target platform will be communicated before running the first
exploit.

@@

Again the name already tells you what this is about, a directed graph that
reflects the control-flow of a program. Different definitions exist regarding
what is actually used to compose a node and anchor them together.In our case we
will create a node out of each function and connect them at function calls.

\begin{figure}[H]
    \centering
    \input{gfx/control_flow_graph.tikz}
    \caption{Control-flow graph example}
    \label{fig:cfg}
\end{figure}

\Cref{fig:cfg} shows an example of such a graph. From this graph you can tell
that \texttt{funB} will be called from \texttt{funA} and \texttt{funE} but not,
for instance, from \texttt{funD}. This is all we care about right now. Note
that contrary to some definitions this graph is not acyclic since we also model
direct and indirect recursion as can be seen by $\texttt{funC} \to
\texttt{funC}$ and $\texttt{funcB} \to \texttt{funcD} \to \texttt{funcE} \to
\texttt{funcB}$ respectively.

@@

CFI is a big topic and, like other topics already mentioned, goes beyond the
scope of this writeup. The first pointer we will hand the reader aims at the
corresponding
section\footnote{\url{http://clang.llvm.org/docs/ControlFlowIntegrity.html}} of
the Clang documentation, but we recommend checking out the related research
paper \cite{ms_cfi} for more information. A more accessible and recent way to
this topic may be the
talk\footnote{\url{https://www.youtube.com/watch?v=FA0VK7s5tSQ}} \textit{New
memory corruption attacks: why can't we have nice things?} given by gannimo
(Mathias Payer).

The CFG has already been established, now let's see how it can be used to
counter the buffer overflow return address dilemma. At compile-time the graph
is available and can be used to create additional constraints which the program
must obey during runtime. This is similar to the StackGuard mechanism where we
attach code to the end of a function which checks if the canary is still
intact. But now we don't check for a canary but for the validity of the return
address directly. From the CFG we can build a set of possible return targets
for each function. Looking back at the example shown in \cref{fig:cfg} we can
determine that \texttt{funB} returns either to \texttt{funA} or \texttt{funE}.
The corresponding addresses are put into this set which is then stored in the
binary. Before \texttt{funB} returns the return address on the stack is
compared to the entries listed in the corresponding set. If no match can be
found, the program terminates.

With this mechanism setup, one can easily see that it is no longer possible to
chain \emph{arbitrary} gadgets together to pull off ROP. But we still control
the return address and can jump to different locations as long as we stick to
the CFG. Each transition from one node to another has to be valid, while the
overall path taking by our ROP chain may do things never intended by the
program's author. This concept is known as control-flow bending.~
\cite{cf_bending}

@@

Using a changed return address is what ultimately enables control-flow bending.
Stack integrity ensures that the same return address is used upon executing the
\texttt{ret} instruction as was pushed upon that very function call. This can
be achieved by using a \textit{shadow-stack} similar to the StackShield
mechanism but we suggest reading about \textit{code-pointer integrity}
\cite{cpi} regarding this topic.

@@

We conclude this section by mentioning the availability of an interpreter
(probably) available in your standard library. As presented by gannimo,
\texttt{printf} is far more capable than just printing arguments. It is also
possible to read and \emph{write} to memory locations. But one can go even
further and craft a format string mimicking each of the eight operators of
Brainfuck\footnote{\url{https://en.wikipedia.org/wiki/Brainfuck}}. Because
Brainfuck is Turing complete we can deduce that \texttt{printf} is a Turing
complete interpreter. Note that this requires \texttt{printf} to be called in a
loop and the format strings may depend heavily on your library's
implementation. (Also not all implementation are Turing complete).

@@

x86\_64, also known as x64 and AMD64, is, at its core, a \SI{64}{\bit}
extension to x86 which already replaced a lot of x86 machines. Its general
purpose registers are \SI{64}{\bit} wide and there are eight more of them.

@@

Contrary, breaking ASLR by brute force gets much harder since the address space
is \emph{much} bigger which yields more entropy for the randomization. Similar
techniques like \textit{heap spraying} basically useless, but we can still
resort to the info leak. Breaking a canary via brute force gets only a little
bit harder, but this has already been shown in an example.

@@

ARM CPUs will be encountered mostly in portable, low-power oriented devices
such as smart phones and tables, but are also found in embedded devices like
routers. They consist of a \SI{32}{\bit} RISC instruction set with a
\SI{16}{\bit} mode (known as THUMB mode). The used calling convention is
basically the same as under x86\_64 (fastcall), arguments are passed via
registers.

Compared to the previous both, ARM has a \SI{4}{\byte} instruction alignment
(\SI{2}{\byte} under THUMB).

@@

Starting with no mitigation mechanisms in place, we have seen how easy it is to
manipulate the program by exploiting just one simple buffer overflow. Going
beyond simple manipulations like changing locale variables, we craft shell code
and injected it to open up a shell accepting and executing arbitrary inputs.
Next, Data Execution Prevention (DEP) was presented to deny the ability of
\emph{injecting} new code.

This was countered by introducing Return Oriented Programming (ROP) (and
ret2libc) which removes the requirement of injecting new code to exploit a
binary. This is done by combining code fragments (gadgets) already available in
the target binary and libraries to build new, malicious sequences of
instructions.

Address Space Layout Randomization (ASLR) was established to prevent ROP but
can be defeated with an information leak. Even the StackGuard mechanism can be
broking with brute force (including an unexpected low upper bound) in certain
scenarios.

The basic idea behind Control-Flow Integrity (CFI) was communicated after that
followed by a small glance at \texttt{printf}'s capabilities to work as an
interpreter. Along the way references and various outlooks have been provided
to aid the reader.

Before concluding with this section a short word about other architectures and
their implications on the presented techniques has been given,
